Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-06-04 large_number_squeezing_with_atoms.tex
Handle Graham's comments, more figure and caption fixes, plus random explanation improvements here and there

At time $t=0$, the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|b\rangle$. As the initial state was a coherent state in both modes, after the population transfer both $|a\rangle$ and $|b\rangle$ are described by coherent states, with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and number number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spatial modes present, result in an overall lowest possible total number variance.

The second problem is due to the number-dependent dynamics of the spatial modes, which is observable even in a semiclassical simulation.  Due to the nonlinear term in the Hamiltonian, the local phase evolution of the atomic field is density dependent.  Consequently the mode shape of the atomic field changes depending on the nonlinearity-density product, which means that unless $U_{aa}|\psi_a({\mathbf{r}})|^2 = U_{bb}|\psi_b({\mathbf{r}})|^2$ for all ${\mathbf{r}}$, the mode shapes will not overlap perfectly during the final coupling pulse. This degraded mode-matching will result in lower efficiency when converting the quadrature squeezing to number squeezing.

@@

\section{Using a $\pi$-pulse to improve mode-matching}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to use a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1}) and (\ref{eqb1}), such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state remains in a single mode (although that mode might change), and exactly half of the atoms are in each state, this technique ensures that both modes have identical mode shapes after the hold time. In practice this will never work perfectly as the mode shapes will change, resulting in non-identical environments for the two modes. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

@@

Fig.~\ref{fig:Combined1Dand2D} shows the non-condensed fraction predicted by the Bogoliubov analysis for one and two dimensions as a function of $r$, which is a scale factor for the size of the trap.  \textbf{Consider not making it a function of r, but simply the size of the trap?}  The traps are assumed to have a constant potential, which minimises the effects of the multimode issues described in section \ref{sec:MMdescription}.  The squeezing in each trap is calculated using the TWA method, and we see that the squeezing does indeed degrade at high $r$ in 1D, but remains constant in $2D$.  For low $r$, the squeezing in 1D is limited by the other multimode effects.  The non-smooth nature of the squeezing prediction in 2D is due to the stochastic nature of the simulations.

In Fig.~\ref{fig:3DBogScaling} we see that the number of accessible states for three-dimensional traps actually decreases with volume.  This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  \textbf{Create these figures (by combining old figures) and check both for latest updates.  Confirm that `simulation' (blue line) means TWA, as opposed to Graham's prediction.}  The non-smooth nature of the Bogoliubov calculation is due to the coarse grid, which leads to granularity as the scale factor is changed.

@@

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/1D_squeezing_scaling.eps}
\caption{1D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig1D_squeezing_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/1D_Bog_mode_occupation_scaling.eps}
\caption{1D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig1D_Bog_mode_occupation_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/2D_squeezing_scaling.eps}
\caption{2D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig2D_squeezing_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/2D_Bog_mode_occupation_scaling.eps}
\caption{2D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig2D_Bog_mode_occupation_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/3D_Bog_mode_occupation_scaling.eps}
\caption{3D experimentalist's scaling. $N=1\times 10^4$, $L=20\mu$m.}
 \label{fig3D_Bog_mode_occupation_scaling}
\end{figure}

>>>>

At time $t=0$, the coupling $\Omega$ is switched on and applied until time $t=t_1$, at which point it is turned off, resulting in a portion of the population being transferred into mode $|b\rangle$. We assume this transfer process fast relative to the nonlinear energy time scale, ensuring that, as the initial states were coherent states, after the population transfer both $|a\rangle$ and $|b\rangle$ are still described by coherent states with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and number number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spatial modes present, result in the overall lowest total number variance. Clearly, however, this averaged number variance will be still be higher than the specific mode that exhibits the best squeezing.

The second problem is due to the number-dependent dynamics of the spatial modes, which is observable even in a semiclassical simulation.  Due to the nonlinear term in the Hamiltonian, the local phase evolution of the atomic field is density dependent.  Consequently the mode shape of the atomic field changes depending on the nonlinearity-density product, which means that unless $U_{aa}|\psi_a({\mathbf{r}})|^2 + U_{ab}|\psi_b({\mathbf{r}})|^2 = U_{bb}|\psi_b({\mathbf{r}})|^2 + U_{ab}|\psi_a({\mathbf{r}})|^2$ for all ${\mathbf{r}}$, the mode shapes will not overlap perfectly during the final coupling pulse. This degraded mode-matching will result in lower efficiency when converting the quadrature squeezing to number squeezing.

@@

\section{Using a $\pi$-pulse to improve mode-matching}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to use a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1}) and (\ref{eqb1}), such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state remains in a single mode, and exactly half of the atoms are in each state, this technique ensures that both modes have identical mode shapes after the hold time. In practice this will never work perfectly --- due to the density distributions of $\psi_a(\mathbf{r})$ and $\psi_a(\mathbf{r})$ not reflecting the ground state, resulting in non-identical environments for the two modes, some modal dynamics will occur. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

@@

Figure \ref{fig1D_squeezing_and_bog_scaling} and Figure \ref{fig2D_squeezing_and_bog_scaling} show the non-condensed fraction predicted by the Bogoliubov analysis compared to that predicted by a full numerical simulation, in one, two and three dimensions, as a function of the trap size. Also shown in the 1D and 2D cases is the maximum amount of number squeezing obtainable, as predicted by the numerical simulation. The full squeezing scaling curve is not shown in 3D, as for the smaller trap sizes the TWA method breaks down, and results are unreliable; for the larger trap sizes with negligible non-condensate fractions, however, the simulation was carried out and a realtive number squeezing of 12.6dB was found. The traps are assumed to have a constant potential, which minimises the effects of the multimode issues described in section \ref{sec:MMdescription}.  We see that the squeezing does indeed degrade at large trap sizes in 1D, but remains constant in $2D$. For small trap sizes, the squeezing in 1D aymptotically approaches that of the ideal 0D system.

We note that the non-smooth nature of the squeezing prediction in 2D is due to the stochastic nature of the simulations, and the necessity of averaging over fewer quantum trajectories due to the very high number grid points in this regime. The non-smooth nature of the non-condensate mode fraction in 3D, on the other hand, is due to the number of modes being both discrete and low in this regime, and counting only the non-condensed fraction in modes with an energy less than than $\mu$, the chemical potential.

In Fig.~\ref{fig:3DBogScaling} we see that the number of accessible states for three-dimensional traps actually decreases with volume in 3D. This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  \textbf{Create these figures (by combining old figures) and check both for latest updates.  Confirm that `simulation' (blue line) means TWA, as opposed to Graham's prediction.}  The non-smooth nature of the Bogoliubov calculation is due to the coarse grid, which leads to granularity as the scale factor is changed.

@@

\begin{figure}
  \centering
  \begin{subfigure}{.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/1D_Bog_mode_occupation_scaling.eps}
  \end{subfigure}
  \begin{subfigure}{.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/1D_squeezing_scaling.eps}
  \end{subfigure}
\caption{1D scaling as function of trap size, $L$. (Left) The proportion the non-condensed fraction in mode $|a\rangle$ relative to total particle number, with the Bogoliubov theory model in red and the full numerical solution in blue. (Right) Relative number squeezing in mode $|a\rangle$ as a function of trap size. Parameters: $N=1\times 10^5$, $U_{aa} = 8.9\times 10^{-40}$Jm, $U_{ab}=U_{bb}=0$, $\tau_{\mathrm{hold}} = 10L$\,s.}
  \label{fig1D_squeezing_and_bog_scaling}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}{.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/2D_Bog_mode_occupation_scaling.eps}
  \end{subfigure}
  \begin{subfigure}{.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/2D_squeezing_scaling.eps}
  \end{subfigure}
\caption{2D scaling as function of trap size, L. (Left) The proportion the non-condensed fraction in mode $|a\rangle$ relative to total particle number, with the Bogoliubov theory model in red and the full numerical solution in blue. (Right) Relative number squeezing in mode $|a\rangle$ as a function of trap size. Parameters: $N=1\times 10^5$, $U_{aa} = 8.9\times 10^{-44}$Jm, $U_{ab}=U_{bb}=0$, $\tau_{\mathrm{hold}} = 10^{5} L^2$\,s.}
  \label{fig2D_squeezing_and_bog_scaling}
\end{figure}

%\begin{figure}
%  \centering
%   \includegraphics[width=7cm]{figures/1D_squeezing_scaling.eps}
%\caption{1D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
% \label{fig1D_squeezing_scaling}
%\end{figure}

%\begin{figure}
%  \centering
%   \includegraphics[width=7cm]{figures/1D_Bog_mode_occupation_scaling.eps}
%\caption{1D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
% \label{fig1D_Bog_mode_occupation_scaling}
%\end{figure}

%\begin{figure}
%  \centering
%   \includegraphics[width=7cm]{figures/2D_squeezing_scaling.eps}
%\caption{2D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
% \label{fig2D_squeezing_scaling}
%\end{figure}

%\begin{figure}
%  \centering
%   \includegraphics[width=7cm]{figures/2D_Bog_mode_occupation_scaling.eps}
%\caption{2D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
% \label{fig2D_Bog_mode_occupation_scaling}
%\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/3D_Bog_mode_occupation_scaling.eps}
\caption{3D scaling as function of trap size, $L$. (Left) The proportion the non-condensed fraction in mode $|a\rangle$ relative to total particle number, with the Bogoliubov theory model in red and the full numerical solution in blue. (Right) Relative number squeezing in mode $|a\rangle$ as a function of trap size. Parameters: $N=1\times 10^4$, $U_{aa} = 4.7\times 10^{-49}$Jm, $U_{ab}=U_{bb}=0$, $\tau_{\mathrm{hold}} = 1.25 \times 10^{11} L^3$\,s.}
 \label{fig3D_Bog_mode_occupation_scaling}
\end{figure}
